Phylogenetics and the human microbiome.

### 2014-04-04 reviews/response.txt

> Reviews from Prof. Frank (Andy) Anderson
> ----------------------------------------

> Thank you for your Systematic Biology submission. It has been reviewed
> by Associate Editor Prof. Olivier Gascuel and three reviewers. Their
> comments are listed at the end of this letter. The reviewers and the
> AE provide some excellent constructive suggestions that I am sure you
> will appreciate.

> I have little to add that wasn't covered by one or more reviewers or
> Olivier in their fine reviews. As an instructor who has almost given
> up on trying to convince microbiology graduate students that they can
> do better analyses than neighbor joining, I was already sensitive to
> some of these differences, but this review helped me appreciate the
> interface between the two disciplines. I agree with Mooers that the
> level of detail varies across the manuscript, but I mainly noticed
> when a topic was summarized quickly and I didn't really understand it
> (I note that Mooers and I both had a question about how tax2tree
> works).

> I have a few comments and questions that may help clarify some of the
> writing. Altogether, I think the manuscript really requires only
> several minor revisions.

> Like Olivier, I note that some of my comments may overlap those of the
> reviewers.

> 1) Pg. 9: "As described above "metagenome" means that data is sheared
> randomly across the genome rather than amplified from a specific
> location," - I assume you mean DNA is sheared randomly?

> 2) Remember, "data" still is a plural noun, at least under my watch!
> Check for "this data", "data is", "data was", etc

> 3) Pg. 10: I don't really understand what "tax2tree" does, based on
> the description here. Does that matter?

> 4) Pg. 14-15: I also don't really understand what UniFrac is or how it
> works, even after looking at Figure 2 (Olivier made a similar
> comment). Could we get some help here?

> 5) The figures need a bit of work. There seems to be some sort of
> equation at the bottom of Figure 1, but I can't decipher it. What do
> the colors mean in Figure 2? I don't find this to be very clear.

> 6) Pg. 20: "How to formulate a useful notion of independent contrasts
> (Felsenstein 1985) on such an object?" - Awkward wording.

> 7) Pg. 23: I don't understand the distinction between "base
> observations" and sample observations here. Can this be clarified?

> 8) Pg. 24: "Future will continue to bring deeper sequencing on more
> samples" - I guess this should be "The future will continue"?

> My recommendation is acceptance pending minor to major changes.
> Actually, no change asked by the reviewers and me (see below) requires
> major efforts, but we all come from different origins, with different
> points of view and expectations, and thus summing all these demands
> will require a bit of time and attention. Basically, we all ask for
> more details and comments here and there, to make the paper more
> precise, comprehensive and self-contained. I am confident that you'll
> find our suggestions useful. I don't believe that the paper should be
> shortened in some places (or very locally), as suggested by Arne
> Mooers. When submitting your revision, please prepare a cover letter
> containing all the comments by the referees and me, with your
> changes/responses.

> Specific comments, possibly redundant with those from the reviewers,
> using Systematic Biology page numbering:

> P4	these techniques AND could

> P8	(Kuczynski et al. 2010). However, for more subtle information

> P8	"for more subtle information deeper sequencing is required" :
> what do you have in mind here?
> Indeed "distinguish between two rather different samples" is a modest
> goal Actually, it's not clear in this subsection what is meant by
> "Microbial community estimation" (title of the subsection). What is
> the goal here? Identify and classify microbes? Measure their
> abundance? Etc.? A few lines and explanations should suffice.

> P11	What is the level of uncertainty in current bacterial
> phylogenies? For example, one commonly considers that only a few
> unknowns remain in mammals phylogeny; what about bacteria? Are people
> discussing the details or the broad lines? Provide a few illustrative
> examples.

> P13 	"There are many OTU inference methods with various speeds and
> strategies" and the two following paragraphs. Describe in a simple way
> the basic principles of these methods, before discussing their
> relative merits (this remark may apply in other places).

> P14 	You describe PD and suggests that it should be used more often.
> But what are the standard alpha diversity measures, and their
> limitations?

> P15 	Are "partially abundance weighted diversity measures" using some
> phylogenetic approaches? Be a bit more specific when describing and
> discussing methods. Here and in many other places (e.g. UniFrac,
> weighted PD, weighted UniFrac, etc.).

> P15 	"Evans and Matsen (2012) showed that weighted UniFrac" check
> sentence.

> P16 	explanations and Fig. 3 on earth-movers distance are unclear to
> me.

> P17 	"By doing so they have recovered more novel protein families
> than they would have using methods organized by taxonomy". I'm lost
> here, as it is said just above that they have derived a
> "phylogeny-driven genomic encyclopedia of Bacteria and Archaea".

> P23-24 Paragraph "In the same vein This would be a false positive." is
> unclear to me.

> P24 Check "The presently used approach is rarefaction, which means
> uniform sub-sampling, typically without replacement to the number of
> reads in the lowest abundance sample"

> P25 Check "Comparative studies will continue to investigate what
> shapes and is shaped by the microbiome"

> P26 "during the 2013 LIRMM Mathematical and Computational Evolutionary
> Biology workshop", change LIRMM into MCEB

> Fig 2. It took time to me to understand and check on the Web that the
> tree must be rooted, to compute the unifrac distance. This is one
> example, among others, that more details are required when describing
> methods, criteria, measures, etc. Some more figures could help, or one
> appendix with formal definitions of the main concepts and tools. At
> least, details and explanations have to be added in the main text.

> Your reference list is extremely long, mostly due to the fact that you
> are citing papers with a lot of authors. Check Syst Biol
> recommendation to authors (I assume that our editorial office will
> answer and help).

> Reviewer 1, Alexis Stamatakis
> -----------------------------

> Comments:
> This is a well written and in depth review on the usage of
> phylogenetics in human microbiome research, which I enjoyed reading.

> I only have a few minor remarks:

> It might be a consideration to cite our recent paper:

> J. Zhang, P. Kapli, P. Pavlidis, A. Stamatakis: "A General Species
> Delimitation Method with Applications to Phylogenetic Placements". In
> Bioinformatics (2013), 29 (22): 2869-2876.

> It would fit will into section "The role of OTUs".

> Apart from that, I believe that a more critical and slightly extended
> discussion of taxonomic classifiers in the very last part of the paper
> should be included. From my point of view, this is still a mostly
> unresolved problem and the currently best classifier, which you may
> also want to cite,  still shows pretty horrible accuracy
> (http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0049334)
> at the genus and family level.

> I would also appreciate if you could explicitly mention the Gartner
> hype curve in the discussion part and provide an estimate as to where
> we are on this curve at present.

> Some typos:

> page 5 lines 7-8 "these techniques and could be used" there seems to
> be a verb missing.

> page 14 lines 12 Here you may add our recent work on PTP, since it
> strives to conduct OTU clustering and phylogenetic inference in one
> single step/pipeline.

> page 15 l 50: in "distance, is special" seems that it should read
> "distance, it is special"

> page 20: lines 51-56: In fact we are currently developing a tool for
> automatic tree extension as described here. While not published yet,
> it is already available as open-source code at:
> http://sco.h-its.org/exelixis/web/software/put/index.html

> Reviewer 2, Arne Mooers
> -----------------------

> This was an interesting manuscript to read, and I compliment whoever
> thought to review Human Microbiome Research (HMR) for the SB
> readership. Overall, there is a lot of very interesting work reported
> on here, though I found the level of detail to be a bit uneven:
> sometimes we are told perhaps too much, and often, what feels like too
> little, in order to make the connections between HMR and SB. There is
> also seems to be some elision between HMR research and metagenomics
> more generally. That isn't a bad thing, but one does sometimes forget
> that we are focussed on HMR.

> Numbers refer to the SB manuscript pagination, rather than to the
> original latex.

> 4-45 additional to
> 5-8 delete and
> 7-23 delete though

> The section "The Human Microbiome" is quite detailed (discussing mouse
> to human transplants wrt obesity, etc.). Given the intense media
> reporting of the work coming from this area, SB readers may not need
> to be primed of the importance of this field. If this were shortened,
> there could be room to expand some of the technical bits that SB
> readers will want that follow. The last part of the section  - about
> sticking to eubacteria and a non-molecular biology approach do seem
> germane.

> 8-12. these data (I guess we will eventually decide that data is
> singular - up to the editors)

> 8-37++ Whereas Woese ++ This is a bit wordy and unnecessary - we know
> that Woese didn't have PCR

> 10-30 Which "databases" ? - this has been written so far as if it were
> being applied to samples, not databases. I am not clear what the unit
> of observation is here.

> 10-32 We quickly switch here from easy background (nonPCR random
> sequencing) to something brand new (at least to me). "At least some
> portion of the rest of the data can be taxonomically classified
> (methods reviewed by Mande et al. 2012); Treangen et al. (2013) report
> speedups and much higher accuracy when reads are assembled before they
> are classified."
> "Rest of data" refers to what - reads found in a database, or reads
> from a sample? "Accuracy" refers to what? "Speed-ups" refer to what?

> 11-53. Here is the first real point of contact between phylogenetics
> and HMR. But what is an "F-measure of precision and recall" - there is
> no citation.

> 12-28 and 12-32. What is the measure of performance that this refers
> to? How do we decide if a method performs well?

> 12-29. Do readers of SB now what "k-mer composition" is? Is it k-means
> clustering?

> 12-37++. Is the MEGAN approach compositional or phylogenetic? I
> believe it is meant to be a hybrid approach (BLAST being
> compositional, but the taxonomic labeling considered phylogenetic),
> but this could be clarified.

> 13-29. It is not clear what heuristics are required for a cut-off
> method. The author says that he is going to describe the OTU inference
> industry, but the paragraph is really just a list of acronyms, and
> then refers obliquely to a "contingency table" of OTU observations.
> So, for example, what is a notion of goodness of clustering that the
> heuristics are trying to achieve? What are the sorts of species
> divisions found in taxonomies? This is an area where more would be
> helpful - and where a shorter introduction might offer space.

> 14-40. Whereas just about every 16S survey investigation involves an
> OTU-based alpha diversity estimate, only a few involve PD: unweighted
> PD has been applied to some 16S survey data (Lozupone and Knight 2007;
> Costello et al. 2009) and to metagenomic reads in a set of marker
> genes (Kembel et al. 2011) - This is an odd pair of sentences: is the
> L&K the only study that used unweighted PD? If so, in what context?

> 14-46. Check this sentence - perhaps start with "While are commonly
> used, abundance weighted PD measures are not.

> 15-30. Perhaps "In contrast to measures of alpha diversity, measures
> of beta diversity do make use of phylogenetics"  This is actually an
> interesting contrast. Is this just historical contingency, or is there
> a logical reason why the two areas disagree on the use of phylogenetic
> information?

> 15-xx. This seems a strong point of contact between metazoan
> phylogenetics and the microbiome folks. UniFrac is clearly described
> on its own terms, except perhaps where I would have written "with the
> fraction of edges connecting samples that are common to two samples"
> rather than "edges that connect between samples" (which doesn't
> strictly make sense). I believe there is the same measure of
> beta-diversity in the community-phylogenetics literature? What is the
> closest analogue?

> 15-4955. The description of Evans and Matsen seems a bit detailed in
> the context -

> 16-16. This requires another sentence of explanation. From this
> reading, the Unifrac method offers a single distance between a pair of
> samples (a ratio of two numbers of edges), rather than a matrix.

> 16-42..51. This is a short paragraph that would seem to explicitly
> link HMR and SB. Why is it so telegraphic?

> 17-45. This seems an odd heading for this section, which seems to be
> about horizontal gene transfer. Given the amount of technical detail
> elsewhere, why are we not told how horizontal gene transfer is
> inferred in these HMR studies?

> Phylogenetic inference. I found this survey very useful. Is there a
> reason this is not an earlier section of the paper, given these are
> the trees upon which many of the comparisons referred to in previous
> sections are based?

> 19-24. "which has ported in the" Is this the correct structure?

> 20-50. "different parameter regimes" - this phrase is used several
> times in this review. What does it mean here? When would we prefer to
> add sequences to a fixed tree rather than re-infer the totality? Is
> there an analogous situation in SB - i.e. is there something SB can
> learn from HMR here?

> I found the challenges section to be rich and well-written, with
> questions that will resonate with the SB community - e.g. the
> discussion of the proxy problem (how edge lengths predict feature
> diversity). The fact that 16S distance does predict whole genome
> distance is highlighted here - it is presented in previous sections,
> but perhaps it could be given more prominence there, given its
> prominence here. This is a general problem.

> 22-10. "Microbial ecology" Perhaps "and explicitly for human microbial
> ecology" to remind us that HMR is the focus here?

> 23-23. "Whole organism approach"  Will this approach be known to SB
> readers? I cannot imagine what is being done here.

> 25-31. "At least for the next several years we can expect more related
> work." What does this mean? Related to what has been reported above?

> 25-32. The future

> I found the final sentence to leave us dangling. After having reviewed
> this manuscript, I could not say exactly where the intersection of SB
> and HMR will be most fruitful for HMR, nor what in HMR might be
> fruitful for SB. The conclusion seems to suggest that phylogenetics
> would not be used in diagnostics, for example. Will it be important in
> primary research then? In constructing the taxonomies the diagnostics
> may require? Finally, the abstract says the review ends by describing
> "the current challenges to phylogenetics coming from this type of
> work."  My reading of the relevant sections is that there is more
> discussion of challenges to phylogenetic HMR research (sampling
> biases, robustness of UniFrac to incorrect tree reconstruction) than
> to phylogenetics per se. The discussion of how to update known trees
> with new sequences is an exception.

> Reviewer 3, Frederic Delsuc
> ---------------------------

> This manuscript presents a review of how phylogenetic tree-thinking
> and methodology is currently used in the study of the human
> microbiome. The author not only reviews the different ways by which
> phylogenetics is currently percolated into microbial ecology, but also
> gives his personal views on the future phylogenetic challenges
> associated with microbiome studies, and insists on the crucial role
> that molecular phylogeneticists has to play in this quickly expanding
> field. Overall, I enjoyed reading this particularly timely manuscript,
> which I find well written. I only have minor comments/suggestions for
> potential improvement.

> Specific comments:
> 1. I think that it should be made clear, from the very beginning of
> the introduction, the definition of the microbiome in order to avoid
> confusion. Indeed, in the current literature, "microbiota" (i.e. the
> collection of microbes living in association with a given organism)
> and "microbiome" (i.e. the collection of genomes of those microbes)
> are more and more often used interchangeably, even though they are not
> referring to the same organizational level. In that sense, the
> definition of the "human microbiome" given at the top of page 5 ("The
> human microbiome is the collection of microbial organisms that live
> inside of and on the surface of humans.") does not seem to be
> accurate. Please clarify.

> 2. In the section introducing the human microbiome, I think that a
> paragraph on the origin and the way in which the microbiome is
> transmitted vertically between the mother and infant is lacking. The
> recent review on maternal transmission by Funkhouser and Bordenstein
> 2013 is cited later, but it would be nice to mention it on page 5,
> especially in the context of the some examples on the impact of birth
> method (vaginal birth vs. cesarean delivery) on future health of
> newborns.

> 3. On page 6, lines 26-27, it might be worth citing also the recent
> paper by Findley et al. (doi:10.1038/nature12171) on the topography of
> fungal and bacterial on human skin.

> 4. On page 6, line 56, maybe also specify that "survey" is the
> equivalent of barcoding, a term with which the Systematic Biology's
> community is more familiar.

> 5. On page 14, lines 14-19, maybe mention the widespread use of
> "rarefaction curves" to enable the comparison between samples of
> different sequencing depth or methods (e.g. 454 vs. Illumina).

> 6. On page 17, lines 54-55, please add some recent references on the
> current debate around the rooting of the placental tree. I don't think
> that most readers of Syst. Biol. will be familiar with this
> discussion.

> 7. On page 18, lines 26-27, RAxML is also implemented in QIIME.

> 8. On page 20, I think that this might be the place to discuss the
> potential use of phylogenetic networks in microbiome research (see
> Parks and Beiko 2012 Mol Biol Evol).

> 9. On page 24, when citing uBiome, it might also be fair to refer the
> similar crowd-funded American Gut Project.

> Typos:
> P3, L40: I would rather use "species delimitation" instead of "species definition".
> P4, L9: "These techniques and could be used"
> P6, L32: "we will be primarily be describing"
> P9, L29, L30: Maybe replace "diversity" by "variability"
> P18, L32: "a long history as a means to sequentially build"
> P19, L16: "GreenGenes 13_5 alignment has is almost 80% gaps"

### 2014-04-05 sbreview.tex
First pass, fixing typos.

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage{wrapfig}
\usepackage[hidelinks]{hyperref}
\usepackage[round]{natbib}
\usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
\newcommand{\forarxiv}[1]{#1}
\newcommand{\notforarxiv}[1]{}

% % for SB submission
% \renewcommand{\section}[1]{%
% \bigskip
% \begin{center}
% \begin{Large}
% \normalfont\scshape #1
% \medskip
% \end{Large}
% \end{center}}
% \renewcommand{\subsection}[1]{%
% \bigskip
% \begin{center}
% \begin{large}
% \normalfont\itshape #1
% \end{large}
% \end{center}}
% \renewcommand{\subsubsection}[1]{%
% \vspace{2ex}
% \noindent
% \textit{#1.}---}
% \renewcommand{\tableofcontents}{}
% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
% \linespread{1.6}

@@

Although there is something of a divide between phylogeny as practiced as part of microbial ecology on one hand and that for multicellular organisms on the other, there are many parallels between the two enterprises.
Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species definition.
However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
The species concept is even more problematic for microbes than for multicellular organisms, and hence there is also considerable discussion concerning how to group them into species-like units.
Organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.

Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for comparing samples rather than comparing on the level of species abundances.
Although there is some overlap with previous literature, these techniques and could be used in a wider setting and may deserve broader consideration by the phylogenetics community.

@@

Bacteria have been the primary focus of human microbiome research, and other domains have been investigated though to a lesser extent.
Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
We will focus on bacteria here.

In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

@@

The Human Microbiome Project \citep{methe2012framework} generated lots of survey, metagenome, and whole-genome sequencing data and this data is available on a dedicated website\footnote{\url{http://www.hmpdacc.org/}}.
The MetaHIT study \citep{qin2010human} also generated lots of data but it is not available to outside researchers.

@@

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the diversity of gene repertoire in microbes, this core gene set may be relatively small.
Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.

Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a special case of the earth-movers distance, is special in that it can be calculated in linear time, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

@@

\subsection{Genome-scale inquiries using phylogenetics}
With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
As described above, whole-genome data is typically used to directly infer functional information rather than information concerning ancestry.
The continuing debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.

@@

\subsection{Alignment and tree inference}
In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
Furthermore, given that the underlying data is typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
Additionally, as specified below, these data sets can be very large.
There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
Given this perspective, it is not surprising that Bayesian phylogenetic methods and methods that incorporate alignment uncertainty are absent.

@@

Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
There are two major projects to do so: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nucleotides,
% SILVA
% 1543 / 42284
% .96353230536373096207
the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
The GreenGenes tree is updated by running FastTree from scratch for every release.
There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.

@@

One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.

@@

It appears that neutral models involving phylogenetics could be more fully developed.
Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
The results of this simple method are reasonable, but would a collection of gene trees reconciled with a species tree allow for better prediction?
Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
Here again the ``tree of life'' problem can be formulated in a practical light: what representation allows for the best prediction of features of underlying genomes?
How to formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
It is quite possible that inference using a more complete representation would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect more related work.
Future will continue to bring deeper sequencing on more samples.
The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

>>>>

\documentclass{amsart}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{palatino}
\usepackage{wrapfig}
\usepackage[hidelinks]{hyperref}
\usepackage[round]{natbib}
\usepackage[labelfont=bf,labelsep=period,font=small]{caption} % bold the 'Figure #' in the caption and separate it with a period
\newcommand{\forarxiv}[1]{}
\newcommand{\notforarxiv}[1]{#1}

% for SB submission
\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}
\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}
\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}
\renewcommand{\tableofcontents}{}
\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
\linespread{1.6}

@@

Although there is something of a divide between phylogeny as practiced as part of microbial ecology on one hand and that for multicellular organisms on the other, there are many parallels between the two enterprises.
Both communities struggle with issues of sequence alignment, large-scale tree reconstruction, and species delimitation.
However, approaches differ between the microbial ecology community and that of eukaryotic phylogenetics, in part because the scope of the former contains an almost unlimited diversity of organisms, leading to additional problems above the usual.
The species concept is even more problematic for microbes than for multicellular organisms, and hence there is also considerable discussion concerning how to group them into species-like units.
Organizing microbes into a sensible taxonomy is a serious challenge, especially in the absence of obvious morphological features.

Because of this high level of diversity and challenges with species definitions, microbial ecology researchers have developed their own explicitly phylogenetic techniques for comparing samples rather than comparing on the level of species abundances.
Although there is some overlap with previous literature, these techniques could be used in a wider setting and may deserve broader consideration by the phylogenetics community.

@@

Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
We will focus on bacteria here.

In this paper we will primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

@@

The Human Microbiome Project \citep{methe2012framework} generated lots of survey, metagenome, and whole-genome sequencing data and these data are available on a dedicated website\footnote{\url{http://www.hmpdacc.org/}}.
The MetaHIT study \citep{qin2010human} also generated lots of data but it is not available to outside researchers.

@@

\subsection{Metagenomes}
As described above ``metagenome" means that DNA is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.

It is possible to use metagenomic data as an expanded set of marker genes.
That is, one can use 16S reads that appear in the metagenome as well as reads from other ``core" genes that are expected to follow the same evolutionary path and are present in a large proportion of micro-organisms.
This is proven to be a useful strategy \citep{von2007quantitative,wu2008amphora,stark2010mltreemap,kembel2011phylogenetic}, however, because of the variability of gene repertoire in microbes, this core gene set may be relatively small.
Indeed, even the largest collection of genes in these databases only recruits around 1 percent of a metagenome.
At least some portion of the rest of the data can be taxonomically classified \citep[methods reviewed by][]{mande2012classification}; \citet{treangen2013metamos} report speedups and much higher accuracy when reads are assembled before they are classified.

Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.

@@

The name UniFrac is a contraction of ``Unique Fraction,'' which refers to the fact that the original UniFrac definition compares the fraction of edges that connect only tips from one sample via the shortest path rather than edges that connect between samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, that it can be calculated in linear time in contrast to many applications of the earth-movers distance, and that the commonly-used randomization procedure to attach a p-value to an observed distance has a central limit theorem approximation as a Gaussian process.
The earth-movers distance in this case can be defined as the minimum amount of work required to move piles of dirt in one configuration to another along the tree; in this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.

@@

\subsection{Genome-scale inquiries using phylogenetics}
With some notable exceptions, mainstream applications of phylogenetics to a collection of human-associated microbes have typically been with the idea of finding ``the'' tree of such a collection rather than explicitly exploring divergence between various gene trees.
As described above, whole-genome data sets are typically used to directly infer functional information rather than information concerning ancestry.
The continuing debate concerning whether a microbial tree of life is a useful concept \citep{bapteste2009prokaryotic,caro2012bacterial} does not seem to have dampened human microbiome researchers' enthusiasm for using a single such tree.

@@

\subsection{Alignment and tree inference}
In general, human microbiome researchers are interested in quickly doing phylogenetic inference on large data sets, and are less interested in clade-level accuracy or measures of uncertainty.
This is defended by saying that for applications such as UniFrac, the tree is used as a framework to structure the data, and there is a certain amount of flexibility in that framework that will give the same results.
Furthermore, given that the underlying data sets are typically 16S alone we can expect some topological inaccuracy in reconstructing the ``tree of cells'' even with the best methods.
Additionally, as specified below, these data sets can be very large.
There does not seem to be contentious discussion of specific features of the inferred trees equivalent to, say, the current discussion around the rooting of the placental mammal tree.
Given this perspective, it is not surprising that Bayesian phylogenetic methods and methods that incorporate alignment uncertainty are absent.

@@

Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
There are two major projects to do so: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nucleotides,
% SILVA
% 1543 / 42284
% .96353230536373096207
the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the 1,262,986 sequence GreenGenes 13\_5 alignment is almost 80\% gap.
The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
The GreenGenes tree is updated by running FastTree from scratch for every release.
There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.

@@

One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
For example, sequence data sets are continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.

@@

It appears that neutral models involving phylogenetics could be more fully developed.
Methods explicitly invoking trait evolution are notably absent, with the recent exception of \citep{langille2013predictive}.
The results of this simple method are reasonable, but would a collection of gene trees reconciled with a species tree allow for better prediction?
Perhaps improved methods, say involving whole-genome evolutionary modeling or models of metabolic network evolution, could shed light on the problem.
Here again the ``tree of life'' problem can be formulated in a practical light: what representation allows for the best prediction of features of underlying genomes?
How might one formulate a useful notion of independent contrasts \citep{felsenstein1985phylogenies} on such an object?
It is quite possible that inference using a more complete representation would not be able to overcome the inherent noise of the data, but further exploration seems warranted as simple methods give reasonable results.

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect more related work.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome project\footnote{\url{http://ubiome.com/}} promises to bring gut microbiome sequencing to the average citizen for a low price \citep{costandi2013citizen}.
Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.
